---
title: 'CKME136 Capstone Project'
author: "Erma Fernando (Student Number 500925471)"
output:
  word_document: default
  pdf_document: default
  
---
# Load Data
### 1. Read the csv files for 2011, 2012, 2013, 1014, 2015
```{r}

data2011<-read.csv(file="C:/RY/CKME136/OntarioGHG data/FinalData/bps_raw_2011_FINAL.csv",header=T,sep=",")
data2012<-read.csv(file="C:/RY/CKME136/OntarioGHG data/FinalData/bps_raw_2012_FINAL.csv",header=T,sep=",")
data2013<-read.csv(file="C:/RY/CKME136/OntarioGHG data/FinalData/bps_raw_2013_FINAL.csv",header=T,sep=",")
data2014<-read.csv(file="C:/RY/CKME136/OntarioGHG data/FinalData/bps_raw_2014_FINAL.csv",header=T,sep=",")
data2015<-read.csv(file="C:/RY/CKME136/OntarioGHG data/FinalData/bps_raw_2015_FINAL.csv",header=T,sep=",")


### 2. Merge the data frames for 2011, 2012, 2013, 1014, 2015
all_data <-rbind(data2011, data2012, data2013, data2014, data2015)

```

# Initial Analysis of the data and Data Preparation

```{r}

# column headers
names(all_data)
# remove columns that are not necessary like Unit of measure etc
all_data<-all_data[,-c(9,10, 17, 18, 19, 22,24,26,28,30,31,32,35,36,37,40,44,45)]
# column headers
names(all_data)
#num of rows,num of columns, data types
str(all_data)
# 88259 rows 27 columns
#convert Year to a factor
all_data$Year<- as.factor(all_data$Year)

# Convert TRUE to 1 and FALSE and NA to 0. Then convert these variables to factors
all_data$SwimmingPool<-ifelse(all_data$SwimmingPool=="TRUE",1,0)

all_data$DistrictHeating_IsRenewable<-ifelse(all_data$DistrictHeating_IsRenewable=="TRUE",1,0)

all_data$DistrictCooling_IsRenewable<-ifelse(all_data$DistrictCooling_IsRenewable=="TRUE",1,0)

all_data$SwimmingPool<- as.factor(all_data$SwimmingPool)

all_data$DistrictHeating_IsRenewable<- as.factor(all_data$DistrictHeating_IsRenewable)

all_data$DistrictCooling_IsRenewable<- as.factor(all_data$DistrictCooling_IsRenewable)

str(all_data)


# How many missing data in columns
sapply(all_data, function(x) sum(is.na(x)))


# replace null values in numerical columns with 0 
all_data[is.na(all_data)] <- 0

# Check rows with missing data - none - because all replaced with 0 in last step
sapply(all_data, function(x) sum(is.na(x)))

#num of rows,num of columns, data types 88615 obs. of  27 variables
str(all_data)
# min, max, median, mean 1st quartile, 3rd quartile
summary(all_data)

# count of various data types -factor 8 numeric 19
table(unlist(lapply(all_data,class)))

#list the unique values in variable "OperationType""
sapply(all_data, function(x) unique(all_data$OperationType))

# Convert operation types to numerical values in new column TypeofOperation

TypeofOperation<-ifelse(all_data$OperationType=='Administrative office',1,
ifelse(all_data$OperationType=='Library',2,   
ifelse(all_data$OperationType=='Water treatment',3, 
ifelse(all_data$OperationType=='Water pumping',4,
ifelse(all_data$OperationType== 'Sewage treatment',5,
ifelse(all_data$OperationType== 'Sewage pumping', 6,
ifelse(all_data$OperationType== 'Police station',7,
ifelse(all_data$OperationType== 'Fire station',8,
ifelse(all_data$OperationType== 'Storage',9,
ifelse(all_data$OperationType== 'Community centres',10,
ifelse(all_data$OperationType== 'Classrooms', 11,
ifelse(all_data$OperationType== 'Hospital', 12,
ifelse(all_data$OperationType== 'Ambulance station', 13,
ifelse(all_data$OperationType== 'Laboratories', 14,
ifelse(all_data$OperationType== 'Student residences', 15,  
ifelse(all_data$OperationType== 'Recreational facilities', 16,  
ifelse(all_data$OperationType== 'School', 17,
ifelse(all_data$OperationType== 'Parking', 18,  
ifelse(all_data$OperationType== 'Indoor swimming pools', 19,
ifelse(all_data$OperationType== 'Indoor ice rinks', 20,
ifelse(all_data$OperationType== 'Multi-use', 21,
ifelse(all_data$OperationType== 'Art galleries', 22,
ifelse(all_data$OperationType== 'Performing arts facilities',  23, 
ifelse(all_data$OperationType== 'Auditoriums', 24,
  ifelse(all_data$OperationType== 'Other', 25 ,26
          )))))))))))))))))))))))))

#convert Operation to a factor
TypeofOperation<- as.factor(TypeofOperation)

#combine the Operation with the dataframe
all_data<-cbind(all_data,TypeofOperation)

#list the unique values in variable Sector
sapply(all_data, function(x) unique(all_data$Sector))


# Convert Sector to numerical values in new column TypeofSector
TypeofSector<-ifelse(all_data$Sector=='Post-Secondary Educational Institution',1,
ifelse(all_data$Sector=='Public Hospital',2,   
 ifelse(all_data$Sector=='Municipal',3, 
  ifelse(all_data$Sector=='School Board',4,5))))

#convert Operation to a factor
TypeofSector<- as.factor(TypeofSector)


#combine the Operation with the dataframe
all_data<-cbind(all_data,TypeofSector)

# 88259 obs. of  29 variables 
str(all_data)

#write back to disk file will contain all 5 years of data if want to verify data
# commented out code below
#write.csv(all_data, file = "C:/RY/CKME136/OntarioGHG data/FinalData/All_Data.csv")


```


# Load the libraries

```{r}
# When Load packages - packages will be stored in C:\software\packages because set env variable R_LIBS_USER

#install.packages("dplyr")
library(plyr)
library(dplyr)
#install.packages("ggplot2")
library(ggplot2)
#install.packages("corrplot")
library(corrplot)


```

# Exploratory phase 
#  Normalize the data (normalize numerical columns)

```{r}

#first remove rows that have value less than 1 KG  for GHG emisssions
#now have 87774 obs. of  29 variables
all_data<-all_data[all_data$GHGEmissions_KG>1,]

str(all_data)
summary(all_data)


# normalize the data
normalize <- function(x){
  return ((x-min(x))/(max(x) - min(x)))
}

all_data_n<- all_data %>%
    mutate_if(is.numeric, normalize)

str(all_data_n)
summary(all_data_n)


```


# Exploratory phase

# Plotting the normalized data to look at correlations and outliers 

```{r}

####### Plots
# plots below show attributes which are correlated to GHG
plot(x=all_data_n$Electricity_kwh, y=all_data_n$GHGEmissions_KG) 
plot(x=all_data_n$NaturalGas_Cubicmeter, y=all_data_n$GHGEmissions_KG) 
plot(x=all_data_n$FuelOil12_L, y=all_data_n$GHGEmissions_KG) 
plot(x=all_data_n$FuelOil46_L, y=all_data_n$GHGEmissions_KG)  

# plots below show attributes which are not correlated to GHG
plot(x=all_data_n$TotalIndoorSpace_sqft, y=all_data_n$GHGEmissions_KG)
plot(x=all_data_n$WeeklyAverageHours, y=all_data_n$GHGEmissions_KG)
plot(x=all_data_n$AnnualFlow_M , y=all_data_n$GHGEmissions_KG)
plot(x=all_data_n$DistrictHeating_GJ , y=all_data_n$GHGEmissions_KG)
plot(x=all_data_n$DistrictCooling_GJ, y=all_data_n$GHGEmissions_KG)
plot(x=all_data_n$EnergyIntensityekWh_sqft, y=all_data_n$GHGEmissions_KG)
plot(x=all_data_n$EnergyIntensity_ekWh_mega_litre, y=all_data_n$GHGEmissions_KG)

# plot of Sector and OPeration type with GHG
plot(x=all_data_n$TypeofSector, y=all_data_n$GHGEmissions_KG, main="GHG across Types of Sectors")  
# Sector 4 Schoolboard has a big spread while other 3 sectors are around 0
plot(x=all_data_n$TypeofOperation, y=all_data_n$GHGEmissions_KG,main="GHG across Types of Operations") 
#TypeofOperation 17 (School) big spread, has greatest GHG followed by 1 (Administrative Office)

######## Boxplots
boxplot(all_data_n$GHGEmissions_KG ~ all_data_n$TypeofOperation, data=all_data_n, main="GHG across Types of Operation")
boxplot(all_data_n$GHGEmissions_KG ~ all_data_n$TypeofSector, data=all_data_n, main="GHG across Sectors")
boxplot(all_data_n$GHGEmissions_KG, horizontal=TRUE,main="GHG emissions")



# GHG and type of sector
plot(all_data_n$Electricity_kwh, all_data_n$GHGEmissions_KG,        # x variable, y variable
     col = all_data_n$TypeofSector,                          # colour by TypeofSector
     pch = 16,                                    # type of point to use
     cex = 2,                                     # size of point to use
     xlab = "Electricity",                       # x axis label
     ylab = "GHG",                       # y axis label
     main = "GHG and Electricity per Type of Sector")     # plot title
legend (x = .8, y = .7, legend = levels(all_data_n$TypeofSector), col = c(1:4), pch = 16)

# GHG and type of Operation
plot(all_data_n$Electricity_kwh, all_data_n$GHGEmissions_KG,        # x variable, y variable
     col = all_data_n$TypeofOperation,                          # colour by TypeofOperation
     pch = 16,                                    # type of point to use
     cex = 2,                                     # size of point to use
     xlab = "Electricity",                       # x axis label
     ylab = "GHG",                       # y axis label
     main = "GHG and Electricity per Type of Operation")     # plot title
legend (x = .9, y = .9, legend = levels(all_data_n$TypeofOperation), col = c(1:26), pch = 16)


# following shows Type of Operation 17 (School) has highest GHG followed by 1 (Administrative Office)
p<-ggplot(all_data_n, aes( all_data_n$GHGEmissions_KG, all_data_n$TypeofOperation))

geom_boxplot(mapping = NULL, data = NULL, stat = "boxplot",
  position = "dodge2",all_data_n, outlier.colour = NULL,
  outlier.color = NULL, outlier.fill = NULL, outlier.shape = 19,
  outlier.size = 1.5, outlier.stroke = 0.5, outlier.alpha = NULL,
  notch = FALSE, notchwidth = 0.5, varwidth = FALSE, na.rm =TRUE,
  show.legend = NA, inherit.aes = TRUE)

p + geom_boxplot()



#  using log scale for GHG - plot GHG versus Type of Operation,blue shows the outliers
ggplot(data = all_data_n, aes(x=all_data_n$TypeofOperation, y=all_data_n$GHGEmissions_KG)) + 
            scale_y_log10() +
            geom_point(aes(color=all_data_n$TypeofOperation), alpha=0.2) +
            geom_boxplot(outlier.size=4, outlier.colour='blue', alpha=0.1)


# using log scale for GHG - plot GHG versus Sector,blue shows the outliers
ggplot(data = all_data_n, aes(x=(all_data_n$TypeofSector), y=all_data_n$GHGEmissions_KG)) + 
            scale_y_log10() +
            geom_point(aes(color=(all_data_n$TypeofSector)), alpha=0.2) +
            geom_boxplot(outlier.size=4, outlier.colour='blue', alpha=0.1)




```

# Exploratory phase 
# Summaries for GHG by Sector and Type of operation


#  Feature optimization - using Principal component analysis (using numerical columns of normalized data excluding GHG)

```{r}
#normalized data - look at spread of values in the variables
#library(dplyr)

#create table of counts of records for each OperationType arrange by descending counts
# Results show that operation type school has the highest numbeof rows
all_data_n %>%
group_by(OperationType) %>%
summarise(freq=n(), totalGHG=sum(GHGEmissions_KG)) %>%
arrange(desc(freq))

#create table of counts of records for each OperationType arrange by descending GHGEmissions.Kg
# Results show that operation type school has the highest TotalGHG
all_data_n %>%
group_by(OperationType) %>%
summarise(freq=n(), totalGHG=sum(GHGEmissions_KG)) %>%
arrange(desc(totalGHG))

#create table of counts of records for each Sector arrange by descending GHGEmissions.Kg
# Results show that sector School Board has the highest numbeof rows
all_data_n %>%
group_by(Sector) %>%
summarise(freq=n(), totalGHG=sum(GHGEmissions_KG)) %>%
arrange(desc(totalGHG))

####### Principal Component Analysis for feature selection

# PCA including GHG
str(all_data_n[,c(9:12,14:21, 23,25:27)])

all_data_n_pca<-prcomp(all_data_n[,c(9:12,14:21, 23,25:27)], scale = TRUE)
summary(all_data_n_pca)
# PCA shows that the variance of first 13 principalcomponents account for 99.9% of the variance
#PC1 has proportion of variance 30.68%, PC2 8.66 %, PC3 7.3% ..
# Principal component analysis may not be the best method for feature selection as it does not narrow down the number of attributes of importance


# PCA excluding GHG column
all_data_n_pca<-prcomp(all_data_n[,c(9:12,14:21, 23,26,27)], scale = TRUE)
summary(all_data_n_pca)
# PCA shows that the variance of first 13 principalcomponents account for 99.9% of the variance]]
#PC1 has proportion of variance 26.1%, PC2 9.2 %, PC3 7.7% ..
# Principal component analysis may not be the best method for feature selection as it does not narrow down the number of attributes of importance

```

# Feature optimization (Dimensionality Reduction) - using the forward selection algorithm

```{r echo=TRUE}

library(MASS)

# all the independent variables are included
full <- lm(all_data_n$GHGEmissions_KG ~ all_data_n$TypeofSector+ all_data_n$TypeofOperation + all_data_n$TotalIndoorSpace_sqft + all_data_n$WeeklyAverageHours + all_data_n$AnnualFlow_M  + all_data$NumberofPortables + all_data_n$Electricity_kwh  + all_data_n$NaturalGas_Cubicmeter  + all_data_n$Propane_Litre  + all_data_n$DistrictHeating_GJ  + all_data_n$DistrictCooling_GJ + all_data_n$FuelOil12_L + all_data_n$FuelOil46_L, data=all_data_n)

# none of the independent variables are selected.
null <- lm(all_data_n$GHGEmissions_KG~1,data=all_data_n)

stepF <- stepAIC(null, scope=list(lower=null, upper=full), direction= "forward", trace=FALSE)

#display results]
summary(stepF)
stepF$anova
###### Results give the following
# Final Model:all_data_n$GHGEmissions_KG ~ all_data_n$FuelOil12_L + all_data_n$NaturalGas_Cubicmeter +     all_data_n$DistrictHeating_GJ + all_data_n$FuelOil46_L +  all_data_n$DistrictCooling_GJ + all_data_n$Electricity_kwh +  all_data_n$TypeofSector

```
```
# Prediction Stage

# Divide the normalized data into training and testing sets 70:30

```{r}

train_index <-sample (1:nrow(all_data_n), 0.7*nrow(all_data_n))
train.set <- all_data_n[train_index,]
test.set <- all_data_n[-train_index,]

str(train.set)
# train.set 61441 obs. of  29 variables
str(test.set)
# test.set 26333 obs. of  29 variables



# Very Good plot - using the training set numerical columns show Correlations between pairs of attributes
# takes very long to run (3 hours)
# shows that GHG is determined by Electricity, Gas, Fuel Oil12 and Fuel Oil45 only
# commented out line below because it takes hours to run

# pairs.panels(train.set[,c(9:12,14:21, 23,25:27)], gap=0, pch=21)
```

# Prediction Stage - build model

# Multivariate Linear Regression model

```{r}

###### Simple multivariate linear regression model
# build model using train.set
###### using lm
set.seed(111)
model_mlr1 <- lm(train.set$GHGEmissions_KG ~ train.set$FuelOil12_L + train.set$NaturalGas_Cubicmeter +     train.set$DistrictHeating_GJ + train.set$FuelOil46_L + train.set$DistrictCooling_GJ + train.set$Electricity_kwh +  train.set$TypeofSector, data=train.set) 
  
summary(model_mlr1)

coefficients(model_mlr1)

aov(model_mlr1)

prediction_1 <- predict(model_mlr1, interval="prediction", newdata =test.set)

# see errors and plot on histogram
errors_1<- prediction_1[,"fit"] - test.set$GHGEmissions_KG
hist(errors_1)

# Compute the root mean square error and find the percentage of cases with less than 25% error.

rmse_1 <- sqrt(sum((prediction_1[,"fit"] - all_data_n$GHGEmissions_KG)^2)/nrow(test.set))

rel_change_1 <- 1 - ((test.set$GHGEmissions_KG - abs(errors_1)) / test.set$GHGEmissions_KG)

pred25_1 <- table(rel_change_1<0.25)["TRUE"] / nrow(test.set) 
paste("RMSE_1:", rmse_1) 
paste("PRED(25):", pred25_1)

###### using lm - dropped variable "TypeofSector"
set.seed(222)
model_mlr2 <- lm(train.set$GHGEmissions_KG ~ train.set$FuelOil12_L + train.set$NaturalGas_Cubicmeter +     train.set$DistrictHeating_GJ + train.set$FuelOil46_L + train.set$DistrictCooling_GJ + train.set$Electricity_kwh, data=train.set) 
  
summary(model_mlr2)

coefficients(model_mlr2)

aov(model_mlr2)

prediction_2 <- predict(model_mlr2, interval="prediction", newdata =test.set)

# see errors and plot on histogram
errors_2<- prediction_2[,"fit"] - test.set$GHGEmissions_KG
hist(errors_2)

# Compute the root mean square error and find the percentage of cases with less than 25% error.

rmse_2 <- sqrt(sum((prediction_2[,"fit"] - all_data_n$GHGEmissions_KG)^2)/nrow(test.set))

rel_change_2 <- 1 - ((test.set$GHGEmissions_KG - abs(errors_2)) / test.set$GHGEmissions_KG)

pred25_2 <- table(rel_change_2<0.25)["TRUE"] / nrow(test.set) 
paste("RMSE_2:", rmse_2) 
paste("PRED(25):", pred25_2)

###### General multivariate linear regression model
###### using glm - dropped variable "TypeofSector"
set.seed(333)
model_mlr3<-glm(train.set$GHGEmissions_KG ~ train.set$FuelOil12_L + train.set$NaturalGas_Cubicmeter +     train.set$DistrictHeating_GJ + train.set$FuelOil46_L + train.set$DistrictCooling_GJ + train.set$Electricity_kwh, data=train.set)

summary(model_mlr3)

coefficients(model_mlr3)

aov(model_mlr3)

#aov gives intercept and coefficient for each variable
# From the result, formula to calculate GHG is


prediction_3 <- predict(model_mlr3, type="response", se.fit=FALSE, newdata =test.set)

# see errors and plot on histogram
errors_3<- prediction_3 - test.set$GHGEmissions_KG
hist(errors_3)

# Compute the root mean square error and find the percentage of cases with less than 25% error.

rmse_3 <- sqrt(sum((prediction_3 - all_data_n$GHGEmissions_KG)^2)/nrow(test.set))

rel_change_3 <- 1 - ((test.set$GHGEmissions_KG - abs(errors_3)) / test.set$GHGEmissions_KG)

pred25_3 <- table(rel_change_3<0.25)["TRUE"] / nrow(test.set) 
paste("RMSE_3:", rmse_3) 
paste("PRED(25):", pred25_3)


aov(model_mlr1)
aov(model_mlr2)
aov(model_mlr3)
paste("RMSE_1:", rmse_1)
paste("RMSE_2:", rmse_2)
paste("RMSE_3:", rmse_3)
coefficients(model_mlr1)
coefficients(model_mlr2)
coefficients(model_mlr3)
paste("PRED(25):", pred25_1)
paste("PRED(25):", pred25_2)
paste("PRED(25):", pred25_3)
```
